Matching the Time of Emergence of Transboundary Fish species to Lead Time for Policy Response Under Climate Change

### 2020-10-15 Scripts/Manuscript.Rmd
Updated figures orders and manuscriopt

---
title: Matching the Time of Emergence of Transboundary Fish species to Lead Time for Policy Response Under Climate Change
subtitle: Challenges to transboundary fisheries under climate change
author: 'Authors: Juliano Palacios-Abrantes*^1^, Thomas Frölicher^2^, Gabriel Reygondeau^13^, U. Rashid Sumaila^1^, Colette Wabnitz^14^, and William W.L. Cheung^1^'
date: 'Affiliations: ^1^Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada. ^2^Climate and Environmental Physics, Physics Institute, University of Bern.'
output:
  word_document: 
    reference_docx: ../References/template.docx
  pdf_document: default
# bibliography: ../References/FishForVisa_Ref.bib
csl: ../References/science.csl
---

@@

# Abstract (150  words)

@@

Climate change shifts in marine species distributions is affecting the current frameworks emplaced to manage species shared fisheries. Here, we rely on a modeling framework with multiple ensemble members to identify the time frame by which shared stocks distribution will change from historical ranges, and the change in the proportion of the distrirbution that each EEZ has under climate change. Our results suggest that 60% of shared species have emerged since 2006 in 87% of the world EEZs. By 2075 all EEZs will have an emerging species and by 2097 and all species will have emerged in at least one EEZ. Average changes in the ratio of species-share between neighboring EEZs will fluctuate between 1 and 10% with some exceeding 25% by 2030 relative to today’s ratio. Our results suggest that most changes will happen before 2030 urging policy to be re-formulated to account for the possible implications of these changes.

# Introduction

Over the last century human activities have fundamentally changed the physicochemical composition of the ocean, warming the waters, decreasing the pH and reducing the oxygen concentration [@IPCC:2019to]. As a consequence, many marine species have changed their distribution [@Poloczanska:2016kk] and will continue to do so in the search of the appropriate environmental conditions needed for their subsistence [@Cheung:2010dt]. Climate change is expected to impact fisheries production worldwide compromising our capacity to reach international sustainability goals such as the SDG 14 - life below water [@UnitedNations2018; @Cheung:2010dt; @Barange:2014kr]. Such impacts can be partly offset with the proper reform of current fisheries management [@Gaines:2018sg; @Cheung:2018dh]. However, these reforms will require comprehensive plans and institutions to manage species that cross international borders, also known as shared stocks [@Gaines:2018sg; @Pinsky:2018cb].

In the early 1980’s the United Nations officialized the concept of Exclusive Economic Zone (EEZs) through the Law of the Sea (UCLOS) [@UnitedNations:1986tl], partitioning the distribution of several species, thus, creating what we know today as shared stocks. As defined by the Food and Agriculture Organization, shared stocks can be classified in three non-exclusive categories. (*i*) Transboundary, those that cross neighboring EEZs; (*ii*) straddling, that, in addition to neighboring EEZs, also visit the adjacent high seas; and (*iii*) highly migratory, mainly tunas and bill-fishes that cover vast oceanic regions [@Munro:2004th]. While countries are responsible for the management of stocks within their EEZs, UNCLOS encourages actions to cooperate when stocks are shared [Article 63, @UnitedNations:1986tl]. Today, an estimate of 633 transboundary fish species are found in different continents and across multiple habitats averaging 48.5 million tones and representing USD 78 billion in fishing revenue between 2005 and 2010 (Chapter 1).

The management, and thus sustainability, of fisheries that target shared species is jeopardized as marine species shift their known distribution to cope with a changing world [@Pinsky:2018cb]. Frequently, the rules placed to manage these species are set to historical trends [@Baudron:2020dj] and do not necessarily consider the biogeography of species [@Alexa], nor the effects of climate change [@FishForVisa]. Misalignment between policies and distributional shifts have previously resulted in unsustainable harvest and international conflict [@Miller:2013iv; @Song:2017va; @Spijkers:2017ij] and are expected to continue towards the end of the $21^{st}$ century. Thus, there is a need to understand the time when climate change will affect the sharing distribution of these species as well as the intensity in order to prepare ocean governance to species on the move [@Pinsky:2018cb; @FishForVisa; @Link:2010ei].

Here we endeavor to address this gap by estimating the Time of Emergence (ToE) [@Hawkins] and Stock Share Ratio (SSR) [@OCPalacios] of 633 transboundary species worldwide. We define ToE as the time when a change  of a particular variable exceeds the historical natural internal variability [@IPCC:2019to], a concept previously used to investigate the emergence of multiple oceanic physical and biogeochemical variables [@Roggers:2015jq; @Keller:2014jv; @Frolicher:2016dg; @Cheung:2020ja]. Moreover, SSR is adopted from Game Theory as a proxy of thread point - the minimum payoff a player is willing to receive in order to cooperate with others [@OCSumaila; @Sumaila:2013].  We rely on a species distribution model driven by output from an Earth system model with ten ensemble members to project the distribution of the 663 selected shared species [@FishForVisa] by 280 EEZs of 198 coastal countries under a high greenhouse gas emission scenario RCP8.5 (See methods). We treated each species in an EEZ as a single stock due to a lack of more detailed spatial data [@FishForVisa] and only considered shared species between neighboring EEZs, rather than the species’ extended distribution (e.g. we did not include the high seas). Our results suggest that many shared species have already emerged and shifted their SSR in most neighboring EEZs posing a threat to the sustainability of shared stocks and the resilience of international treaties.

# Materials and methods

## Databases and species selection

We included in this analysis 633 marine exploited transboundary species representing 80% of the catch taken from the world's EEZs between 2005 and 2014 [@FishForVisa]. Each species within an EEZ was considered a single stock unit due to a lack of more detailed spatial information [@Teh:2015gd; @FishForVisa]. We defined the boundaries of the world EEZs (e.g., Mexico Pacific EEZ and Mexico Atlantic EEZ) using the *Sea Around Us* shapefile (updated 1 July 2015, available from http://www.seaaroundus.org) noting that it subdivides the EEZs of 198 coastal states into 280 regions (e.g., Mexico's EEZ is divided in Mexico Pacific and Mexico Atlantic; Fig. S1) including islands territories, and determined the intersections between polygons using the R package *sf* [@PackagesfSimpleF:2018vp]. Each EEZ was attributed to a geopolitical region according to the UN and biome [@Reygondeau:2019uh], and species were attributed a habitat preference [@FishBase]. For each species and EEZ we used the *Sea Around Us* data to estimate the catch and fishing revenue from fishing activities within the global EEZs. We report both catch and revenue the average from the last decade (2005-2014) of available spatial data [@Zeller2016; @Tai2017; @Sumaila:2015uc]. The monthly average US Consumer Price Index (CPI) according to the U.S. Bureau of Labor Statistic (https://www.bls.gov/cpi/) was used to convert the original 2010 real USD value to 2019 real USD.

## Projecting species distributions under climate change

We projected the distribution of marine species using a dynamic bioclimatic envelope model (hereafter called DBEM). The DBEM algorithm includes species physiology, habitat suitability, depth and latitudinal ranges, and spatial population dynamics as well as preferences to sea temperature, salinity, oxygen content, sea ice extent (for polar species) and bathymetry. For pelagic species, the model uses environmental variables at the surface (e.g. sea surface temperature) whereas demersal species distribution are driven by ocean-bottom variables (e.g. bottom temperature). This way, the model estimates species abundance and maximum catch potential (a proxy of maximum sustainable yield) over a $0.5^{\circ}$ x $0.5^{\circ}$ grid of the world (See Cheung *et al.* [-@Cheung:2010dt; -@Cheung:2016jd] for model details).

We coupled the DBEM with the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project species distribution from 1951 to 2100 under climate change [@Dunne:2012bo; @Dunne:2013db]. We run the DBEM simulating biomass and MCP from 1950 to 2100 under a “worst-case” climate change scenario according to the Representative Concentration Pathway 8.5 (RCP 8.5) used by the Intergovernmental Panel on Climate Change (IPCC) [@IPCC:2014bz; Riahi:2011dk]. According to the GFDL-ESM2M, under the RCP 8.5 scenario atmospheric temperature will increase by $3.2^{\circ}C$ by 2081-2100, relative to preindustrial levels. Because the main approach of this paper relies on understanding the spatial and temporal variation of a species distribution, we have to understand both the historical as well as the future climate-driven distribution variability, in order to infer the differences. We do this by following an ensemble member approach where each member constitutes a different initial Earth system condition [@Rodggers:2015jq; @Frolicher:2009ge]. Thus, for our experiment, in each of the ten GFDL-ESM2M-ensembles the Earth system initial condition is infinitesimally different resulting in a unique atmosphere and ocean state at each point in time. This results in a different state of internal variability for each ensemble representing a plausible future under RCP 8.5 [@Rodggers:2015jq]. Following this approach allows us to estimate a secular trend for the natural variability (e.g. background noise) and isolate the climate signal of a stock’s distribution by averaging the secular trend over all ensemble members.

## Estimating the Time of Emergence of transboundary species

Knowing the time (e.g., year) in which a shared stock is going to diverge from a historical variability is key for preparing to shifting species [@Link:2010ei; @FishForVisa]. Here, we coupled the DBEM with the ten GFDL ensemble members described above, and developed an index of transboundary range shift, fisheries and climate change (*TI*). The index is based on the variation of a species distribution centroid in relation to the centroid of the neighboring EEZs it occupies. Each species centroid was determined by the average latitude and longitude of the grid cells containing the top 10% abundance percentile of each year derived from the DBEM. The EEZ centroid was estimated using the *st* package of *R* statistical software (**Supplements Fig 1 (SuppM)**). For each ensemble member, neighboring EEZ and species, we computed the distance between centroids assuming a perfect sphere and ignoring elevation differences using the *geosphere* package of *R* statistical software;

@@

where $D_A$ and $D_B$ represent the distance between EEZ and distribution centroids of a species in EEZs A and B, respectively for each future time step from 2006 to 2100 (*t*); *sd* is the standard deviation of the historical (*th*, 1951 - 2005) centroid distribution for $D_A$ and $D_B$. The index was then smoothed to a 10 years average.  Finally, we determined the time of emergence setting an arbitrary threshold following Rodggers (2015):

@@

where *TREND* is the yearly mean *TI* ($\mu$) of all ten ensemble members and *NOISE* is the standard deviation of the TREND ($\sigma$) between 1951 and 2005. 
<!-- Note that in some cases not all ensemble members show the presence of a species in a certain EEZ throughout the study's time frame. We presented the results only for those cases where all 10 ensemble members had results for each year between 1951 and 2100 (n = 150). -->
We estimated the mean and one standar deviation ($\mu \pm \sigma$ for a confidence interval of 68%) to show the first year when the "signal" of the transboundary index overpasses the historic "noise". Since this was an arbitrary decision, we have included a more conservative estimation ($\mu \pm 2*\sigma$ for a 95% confidence interval). The SNR for both bottom and surface sea temperature was also estimated for statistical analysis following the SNR method.

### Vulnerability index

@@

## Changes in transboundary Species Share Ratio

We adopted the concept of threat point as the minimum proportion of the species’ joint distribution between neighboring EEZs that each country will accept if they were to cooperate in the management of the stock [@OCPalacios]. For each species and time step (1951-2100) we first aggregate the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells where the species was present within neighboring EEZs, and then calculated the proportion that each EEZ had [@FishForVisa]. We then average the results in three time periods in order to reduce climate variability within the projections. The first time period ($t_h$) comprises (1951 - 2005) and is assumed as the historic share distribution of each species, thus, representing each neighboring nation's threat point. This time frame matches the historic climate data that the GFDL uses to project climate change [@Dunne:2012bo; @Dunne:2013db]. We then selected two future periods; early $21^{st}$ century as the average of 2021 to 2040 ($t_e$), and mid $21^{st}$ century as the average of 2041 to 2060 ($t_m$). We choose these time periods to match the challenges of achieving fisheries-related UN-SDGs [@UnitedNations2018] such as SDG 14.4 (end overfishing) and SDG 1.2 (poverty reduction), aimed to be achieved by 2020 and 2030, respectively [@Singh:2017ds]. The analysis was replicated for each of the ten ensemble members capturing the variation by averaging results among ensemble members ($\mu \pm \sigma$). We then estimate the percentage change in the SSR ($\Delta{SSR}$) of each future time period ($SSR_{f}$) relative to the historic time period ($SSR_h$) as follows,

@@

## Time of Emergence of transbounudary species

We estimated the ToE between neighboring EEZs for 633 transboundary species, that is, the year where the future stock's distribution (2006-2100) overpasses the historical variation (1951-2005). Since 2006, 87% of the world EEZs have already seen at least one emerging stock and 60% of the transboundary species analyzed have emerged in at least a set of neighboring EEZ (e.g. *Engraulis ringens* shared by Chile and Peru emerged in 2010). species will continue to emerge steadily towards 2097 when all 633 species will have emerged in at least one neighboring EEZ (Fig. 1). The number of EEZs with emerging species will increase steeply until 2020 when it seems to flatten approaching 2081 when all of the EEZs analyzed here will have at least one emerging stock (Fig. 1).

```{r Figure1, eval = T,  echo = F, fig.width = 5, fig.height=5, fig.cap =  "Cumulative increase in time of emergence by Exclusive Economic Zone and transboundary species in absolute numbers and percentage."}

@@

Most of the world EEZ’s will see an average ToE across transboundary species before 2040 (Fig. 2). In a few EEZs species will have a relative late average ToE (e.g. Honduras average ToE is 2071) while 54 of the 280 EEZs (19%) have already seen their species emerged (e.g. average ToE across all species <= 2020). A higher number of emerging species will happen between latitudes 0$^o$ and 50$^o$N corresponding to regions of relatively small EEZs with multiple borders (e.g. The Caribbean and the Mediterranean) (Fig.2). In most cases, the current participation of the emerging species in a country's total fishing revenue is less than 50% (Fig. 2 - Polygons). However, in some cases emerging species represent 85% of current transboundary fishing revenue (e.g. Peru) or more (e.g. Marshall Islands, 92%). Only in very few cases like Brazil and Alaska's polar EEZ, no specie will emerge between 2006 and 2100.

```{r Figure2, eval = T,  echo = F, fig.width  = 12, fig.height =10, fig.cap =  "Average time of emergence of transboundary specie. Land polygon shows the vulnerability index. Exclusive Economic Zone polygon represents the average time of emergence across all species. Density plot on the right shows the latitudinal number of emerging species across EEZ centroids (10^o running mean)."}

@@

Twelve of the 47 countries responsible for 75% of the yearly revenue generated from transboundary species between 2005 and 2010 will see an average ToE before 2030 (Fig. 3). This includes the United States who occupies the $4^{th}$ place in revenue generated from transboundary fisheries. These 47 countries are also among those with the largest number of emerging taxa. Regionally speaking, most of African, Asian and European EEZs will see a latter ToE, while the Americas and Oceania will see an early ToE (Fig. 3). From the top ten countries with more revenue from transboundary species, nine of them share a border. Most of these countries (e.g. China) will also have a large number of emerging species, while others like Peru and Norway will have a small number of emerging taxa but representing high revenue fisheries.

```{r Figure3, eval = T,  echo = F, fig.width=5, fig.height=6, fig.cap =  "Average time of emergence of transboundary species per fishing entity. Fishing revenue in logarithmic scale. Showing names for the top 75th percentile. Countries marked with an asterisk have at least one neighboring border in the group."}

@@

We compared species ToE within EEZs by geographic region and by species habitat preference (Fig. 4). The median ToE of transboundary species varied significantly according to the geographic region of the neighboring EEZs (Kruskal-Wallis, $X^2$ = 287.23, *DF* = 93, *p* < 0.001). Post-hoc tests revealed regional patterns. EEZs in Easter Asia which have the latest average ToE across all habitat types (2041), and Latin America and the Caribbean and Polynesia who have the earliest average ToE across all habitats (2021 and 2018, respectively), significantly differ from (almost) all other regions. Other regions also significantly differ from each other but in a less frequent manner (See Table S1 for full statistics). When looking specifically at habitat association, we found significant differences in ToE by habitat preference (Kruskal-Wallis, $X^2$ = 286.48, *DF* = 93, *p* < 0.001). Specifically, the ToE of species with preferences for pelagic-oceanic and reef habitats was significantly different than other habitat preferences (See Table S2 for full statistics).

```{r Figure4, eval = T,  echo = F, fig.width=14, fig.height=14, fig.cap =  " Regional time of emergence per habitat preference. Regions according to the United Nations and habitat preference to FishBase. Whiskers represent 1.5* interquartile range. Box represents interquartile range as distance between first and third quartiles. Line represents median, and black points represent outliers (outside of 1.5*IQR."}

@@

## Changes in specie Share Ratio (SSR)

```{r analysis_results, eval = T,  echo = FALSE, include = FALSE}

@@

# Discussion

In this paper we endeavor to look at the timing and intensity of changes in the sharing of transboundary species. Our findings here are noteworthy in several ways. First, we show that the current distribution of most transboundary species have already emerged from the historical record and that virtually all EEZs have seen at least one emerging species since 2006 (Fig. 5). Second, we identify regional patterns of complex changes in SSR and levels of vulnerability of transboundary fisheries management to climate change (Fig. 3). Third, we identified that the historical share proportion of transboundary species will change by as soon as 2030 in most neighboring EEZs (Fig. 5).

### Emergence characteristics of transboundary species

>>>>

---
title: Matching the Time of Emergence of Transboundary Fish species to Lead Time for Policy Response Under Climate Change
subtitle: Challenges to transboundary fisheries under climate change
author: 'Authors: Juliano Palacios-Abrantes*^1^, Thomas Frölicher^2^, Gabriel Reygondeau^13^, U. Rashid Sumaila^1^, Colette Wabnitz^14^, and William W.L. Cheung^1^'
date: 'Affiliations: ^1^Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada. ^2^Climate and Environmental Physics, Physics Institute, University of Bern.'
output:
  word_document: 
    reference_docx: ../References/template.docx
  pdf_document: default
bibliography: ../References/ToE_Ref.bib
csl: ../References/science.csl
---

@@

<!-- # Abstract (150  words) -->

@@

<!-- Climate change shifts in marine species distributions is affecting the current frameworks emplaced to manage species shared fisheries. Here, we rely on a modeling framework with multiple ensemble members to identify the time frame by which shared stocks distribution will change from historical ranges, and the change in the proportion of the distrirbution that each EEZ has under climate change. Our results suggest that 60% of shared species have emerged since 2006 in 87% of the world EEZs. By 2075 all EEZs will have an emerging species and by 2097 and all species will have emerged in at least one EEZ. Average changes in the ratio of species-share between neighboring EEZs will fluctuate between 1 and 10% with some exceeding 25% by 2030 relative to today’s ratio. Our results suggest that most changes will happen before 2030 urging policy to be re-formulated to account for the possible implications of these changes.  -->

## Introduction

Over the last century human activities have fundamentally changed the physicochemical composition of the ocean, warming the waters, decreasing the pH and reducing the oxygen concentration [@IPCC:2019tn]. As a consequence, many marine species have changed their distribution [@Poloczanska:2016kk] and will continue to do so in the search of the appropriate environmental conditions needed for their subsistence [@Cheung:2010dt]. Climate change is expected to impact fisheries production worldwide compromising our capacity to reach international sustainability goals such as the SDG 14 - life below water [@UnitedNations2018; @Cheung:2010dt; @Barange:2014kr]. Such impacts can be partly offset with the proper reform of current fisheries management [@Gaines:2018sg; @Cheung:2018dh]. However, these reforms will require comprehensive plans and institutions to manage species that cross international borders, also known as shared stocks [@Gaines:2018sg; @Pinsky:2018cb].

In the early 1980’s the United Nations officialized the concept of Exclusive Economic Zone (EEZs) through the Law of the Sea (UCLOS) [@UnitedNations:1986tl], partitioning the distribution of several species, thus, creating what we know today as shared stocks. As defined by the Food and Agriculture Organization, shared stocks can be classified in three non-exclusive categories. (*i*) Transboundary, those that cross neighboring EEZs; (*ii*) straddling, that, in addition to neighboring EEZs, also visit the adjacent high seas; and (*iii*) highly migratory, mainly tunas and bill-fishes that cover vast oceanic regions [@Munro:2004th]. While countries are responsible for the management of stocks within their EEZs, UNCLOS encourages actions to cooperate when stocks are shared [Article 63, @UnitedNations:1986tl]. Today, an estimate of 633 transboundary fish species are found in different continents and across multiple habitats averaging 48.5 million tones and representing USD 78 billion in fishing revenue between 2005 and 2010 (Chapter Two).

The management, and thus sustainability, of fisheries that target shared species is jeopardized as marine species shift their known distribution to cope with a changing world [@Pinsky:2018cb]. Frequently, the rules placed to manage these species are set to historical trends [@Baudron:2020dj] and do not necessarily consider the biogeography of species [@FredstonHermann:2018kp], nor the effects of climate change (Chpater One). Misalignment between policies and distributional shifts have previously resulted in unsustainable harvest and international conflict [@Miller:2013iv; @Song:2017va; @Spijkers:2017ij] and are expected to continue towards the end of the $21^{st}$ century. Thus, there is a need to understand the time when climate change will affect the sharing distribution of these species as well as the intensity in order to prepare ocean governance to species on the move [@Pinsky:2018cb; @Link:2010ei, Chpater One].

Here we endeavor to address this gap by estimating the Time of Emergence (ToE) [@Hawkins:2012hc] and Stock Share Ratio (SSR) (Chapter Four) of 633 transboundary species worldwide. We define ToE as the time when a change  of a particular variable exceeds the historical natural internal variability [@IPCC:2019tn], a concept previously used to investigate the emergence of multiple oceanic physical and biogeochemical variables [@Rodgers:2015jq; @Keller:2014jv; @Frolicher:2016dg; @Cheung:2020ja]. Moreover, SSR is adopted from Game Theory as a proxy of thread point - the minimum payoff a player is willing to receive in order to cooperate with others [@Sumaila:2020tp; @Sumaila:2013hv].  We rely on a species distribution model driven by output from an Earth system model with ten ensemble members to project the distribution of the 663 selected shared species (Chpater One) by 280 EEZs of 198 coastal countries under a high greenhouse gas emission scenario RCP8.5 (See methods). We treated each species in an EEZ as a single stock due to a lack of more detailed spatial data (Chpater One) and only considered shared species between neighboring EEZs, rather than the species’ extended distribution (e.g. we did not include the high seas). Our results suggest that many shared species have already emerged and shifted their SSR in most neighboring EEZs posing a threat to the sustainability of shared stocks and the resilience of international treaties.

## Materials and methods

### Databases and species selection

We included in this analysis 633 marine exploited transboundary species previously identified by representing 80% of the catch taken from the world’s EEZs between 2005 and 2014 (Chapter Two). Each species within an EEZ was considered a single stock unit due to a lack of more detailed spatial information per exploited population [@Teh:2015gd, Chpater One]. We defined the boundaries of the world EEZs using the *Sea Around Us* spatial division (updated 1 July 2015, available from http://www.seaaroundus.org) noting that it subdivides the EEZs of 198 coastal states into 280 regions (e.g., Mexico’s EEZ is divided in Mexico Pacific and Mexico Atlantic; Figure \@ref(fig:FigureS1)) including islands territories, and determined the intersections between polygons using the R package sf [@PackagesfSimpleF:2018vp]. Each EEZ was attributed to a geopolitical region according to the UN and biome [@Reygondeau:2019uh], and species were attributed a habitat preference according to FishBase. For each species and EEZ we used the *Sea Around Us* data to estimate the catch and fishing revenue from fishing activities within the global EEZs. We report both catch and revenue the average from the last decade (2005-2014) of available spatial data [@Zeller2016; @Tai2017; @Sumaila:2015uc]. The monthly average US Consumer Price Index (CPI) according to the U.S. Bureau of Labor Statistic (https://www.bls.gov/cpi/) was used to convert the original 2010 real USD value to 2019 real USD.

### Projecting species distributions under climate change

We projected the distribution of marine species using a dynamic bioclimatic envelope model (hereafter called DBEM) [@Cheung:2010dt; @Cheung:2016jd]. The DBEM algorithm includes species physiology, habitat suitability, depth and latitudinal ranges, and spatial population dynamics as well as preferences to sea temperature, salinity, oxygen content, sea ice extent (for polar species) and bathymetry. For pelagic species, the model uses environmental variables at the surface (e.g. sea surface temperature) whereas demersal species distribution are driven by ocean-bottom variables (e.g. bottom temperature). This way, the model estimates species abundance and maximum catch potential (a proxy of maximum sustainable yield) over a $0.5^{\circ}$ x $0.5^{\circ}$ grid of the world (See Cheung *et al.* [-@Cheung:2010dt; -@Cheung:2016jd] for model details).

We forced the DBEM with environmental data from a ten ensemble member simulation of  the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project species distribution from 1951 to 2100 [@Dunne:2012bo; @Dunne:2013db]. The GFDL ESM2M is run under historical forcing till 2005 and follows the high ” climate change emission scenario, the Representative Concentration Pathway 8.5 (RCP 8.5) over the 2006-2100 period [@Riahi:2011dk]. In GFDL-ESM2M, the atmospheric surface temperature is projected to increase by $3.2^{\circ}C$ by 2081-2100, relative to preindustrial levels. Because the main approach of this paper relies on understanding the spatial and temporal variation of a species distribution, we have to understand both the historical as well as the future climate-driven distribution variability, in order to infer the differences. We do this by following an ensemble member approach where each member constitutes a different realization of the Earth system variability condition [@Rodgers:2015jq; @Frolicher:2009ge]. Thus, for our experiment, each of the ten GFDL-ESM2M-ensembles were started from infinitesimally different Earth system initial conditions resulting in a unique atmosphere and ocean state at each point in time. Following this approach allows us to estimate the natural internal variability (e.g. background noise) and isolate the forced climate signal of a stock’s distribution by averaging the secular trend over all ten ensemble members.

### Estimating the Time of Emergence of transboundary species

Knowing the time (e.g. year) in which a shared stock is going to diverge from the natural internal variability is key for preparing to shifting stocks [@Link:2010ei]. Here, we coupled the DBEM with the ten GFDL ensemble members described above, and developed an index of transboundary range shift, fisheries and climate change (TI). The index is based on the variation of a species distribution centroid in relation to the centroid of the neighboring EEZs it occupies. Each species centroid was determined by the average latitude and longitude of the grid cells containing the top 10% abundance percentile of each year derived from the DBEM. The EEZ centroid was estimated using the st package of R statistical software (Figure \ref{fig:FigureS1}). For each ensemble member, neighboring EEZ and species, we computed the distance between centroids assuming a perfect sphere and ignoring geographic barriers using the geosphere package of R statistical software;

@@

where $D_A$ and $D_B$ represent the distance between EEZ and distribution centroids of a species in EEZs A and B, respectively for each time step from 2006 to 2100 (t); sd is the standard deviation of the historical (*th*, 1951 - 2005) centroid distribution for $D_A$ and $D_B$. The index was then smoothed to a 10 years average. Finally, we determined the time of emergence following Rodgers et al. (2015):

@@

where *TREND* is the yearly mean *TI* ($\mu$) of all ten ensemble members and *NOISE* is the standard deviation of the TREND ($\sigma$) between 1951 and 2005. 
<!-- Note that in some cases not all ensemble members show the presence of a species in a certain EEZ throughout the study's time frame. We presented the results only for those cases where all 10 ensemble members had results for each year between 1951 and 2100 (n = 150). -->

We estimated the mean and one standard deviation ($\mu \pm \sigma$ for a confidence interval of 68%) to show the first year when the “signal” of the transboundary index overpasses the historic “noise”. Since this was an arbitrary decision, we have included also a more conservative estimation in the supplements ($\mu \pm 2*\sigma$ for a 95% confidence interval). The ToE for both bottom and surface sea temperature was also estimated for statistical analysis following the SNR method.

#### Vulnerability index

@@

### Changes in transboundary Species Share Ratio

We adopted the concept of threat point as the minimum proportion of the species’ joint distribution between neighboring EEZs that each country will accept if they were to cooperate in the management of the stock (Chapter Four). For each species and time step (1951-2100) we first aggregate the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells where the species was present within neighboring EEZs, and then calculated the proportion that each EEZ had Chpater One. We then average the results in three time periods in order to reduce climate variability within the projections. The first time period ($t_h$) comprises (1951 - 2005) and is assumed as the historic share distribution of each species, thus, representing each neighboring nation's threat point. This time frame matches the historic climate data that the GFDL uses to project climate change [@Dunne:2012bo; @Dunne:2013db]. We then selected two future periods; early $21^{st}$ century as the average of 2021 to 2040 ($t_e$), and mid $21^{st}$ century as the average of 2041 to 2060 ($t_m$). We choose these time periods to match the challenges of achieving fisheries-related UN-SDGs [@UnitedNations2018] such as SDG 14.4 (end overfishing) and SDG 1.2 (poverty reduction), aimed to be achieved by 2020 and 2030, respectively [@Singh:2017ds]. The analysis was replicated for each of the ten ensemble members capturing the variation by averaging results among ensemble members ($\mu \pm \sigma$). We then estimate the percentage change in the SSR ($\Delta{SSR}$) of each future time period ($SSR_{f}$) relative to the historic time period ($SSR_h$) as follows,

@@

### Time of Emergence of transbounudary species

We estimated the ToE between neighboring EEZs for 633 transboundary stocks, that is, the first year where the future stock’s distribution (2006-2100) overpasses the historical variation (1951-2005). Since 2006 till 2020, 87% of the world EEZs have already seen at least one emerging stock and 60% of the transboundary species analyzed have emerged in at least a set of neighboring EEZ (e.g., *Engraulis ringens* shared by Chile and Peru emerged in 2010). Stocks will continue to emerge steadily towards 2097 when all 633 species will have emerged in at least one neighboring EEZ (Figure \ref{fig:Figure1}). The number of EEZs with emerging stocks will increase steeply until 2025 when it seems to flatten approaching 2081 when all of the EEZs analyzed here will have at least one emerging stock (Figure \ref{fig:Figure1}).

```{r Figure1, eval = T,  echo = F, fig.width = 9, fig.height=4, fig.cap =  "Cumulative increase in time of emergence by Exclusive Economic Zone and transboundary species in absolute numbers and percentage."}

@@

Most of the world EEZs are projected to experience an average ToE across transboundary species before 2040 (Figure \ref{fig:Figure2}). In a few EEZs species are projected to have a relative late average ToE (e.g. Honduras average ToE is 2071) while 54 of the 280 EEZs (19%) have already seen their species emerged (e.g. average ToE across all species $\leq$ 2020). A higher number of emerging species will happen between latitudes 0$^o$ and 50$^o$N corresponding to regions of relatively small EEZs with multiple borders (e.g. The Caribbean and the Mediterranean) (Figure \ref{fig:Figure2}). In most cases, the current participation of the emerging species in a country’s total fishing revenue is less than 50% (Figure \ref{fig:Figure2}- Polygons). However, in some cases emerging species represent 85% of current transboundary fishing revenue (e.g. Peru) or more (e.g., Marshall Islands, 92%). Only in very few cases like Brazil and Alaska’s polar EEZ, no stock will emerge between 2006 and 2100.

```{r Figure2, eval = T,  echo = F, fig.width  = 13, fig.height =11, fig.cap =  "Average time of emergence of transboundary specie. Land polygon shows the vulnerability index. Exclusive Economic Zone polygon represents the average time of emergence across all species. Density plot on the right shows the latitudinal number of emerging species across EEZ centroids (10^o running mean)."}

@@

Regionally, most of African, Asian and European EEZs are projected to have a later ToE, while the Americas and Oceania will see an early ToE (Figure \ref{fig:Figure3}). Twelve of the 47 countries responsible for 75% of the yearly revenue generated from transboundary species between 2005 and 2010 will see an average ToE before 2030 (Figure \ref{fig:Figure3}). This includes the United States who occupies the $4^{th}$ place in revenue generated from transboundary fisheries. These 47 countries are also among those with the largest number of emerging taxa (Figure \ref{fig:Figure3}). From the top ten countries with more revenue from transboundary species, nine of them share a border. Most of these countries (e.g. China) will also have a large number of emerging stocks, while others like Peru and Norway will have a small number of emerging taxa but representing high revenue fisheries.

```{r Figure3, eval = T,  echo = F, fig.width=4, fig.height=5, fig.cap =  "Average time of emergence of transboundary species per fishing entity. Fishing revenue in logarithmic scale. Showing names for the top 75th percentile. Countries marked with an asterisk have at least one neighboring border in the group."}

@@

We compared species ToE within EEZs by geographic region and by species habitat preference (Figure \ref{fig:Figure4}). The median ToE of transboundary species varied significantly according to the geographic region of the neighboring EEZs (Kruskal-Wallis, $X^2$ = 287.23, *DF* = 93, *p* < 0.001). Post-hoc tests revealed regional patterns. EEZs in Easter Asia which have the latest average ToE across all habitat types (2041), and Latin America and the Caribbean and Polynesia who have the earliest average ToE across all habitats (2021 and 2018, respectively), significantly differ from (almost) all other regions. Other regions also significantly differ from each other but in a less frequent manner (see Table \ref{tab:TableS1}). When looking specifically at habitat association, we found significant differences in ToE by habitat preference (Kruskal-Wallis, $X^2$ = 286.48, *DF* = 93, *p* < 0.001). Specifically, the ToE of species with preferences for pelagic-oceanic and reef habitats was significantly different than other habitat preferences (See Table \ref{tab:TableS2}).

```{r Figure4, eval = T,  echo = F, fig.width=14, fig.height=14, fig.cap =  "Regional time of emergence per habitat preference. Regions according to the United Nations and habitat preference to FishBase. Whiskers represent 1.5* interquartile range. Box represents interquartile range as distance between first and third quartiles. Line represents median, and black points represent outliers outside of 1.5*IQR."}

@@

### Changes in specie Share Ratio (SSR)

```{r analysis_results, eval = F,  echo = FALSE, include = FALSE}

@@

## Discussion

In this paper we endeavor to look at the timing and intensity of changes in the sharing of transboundary species. Our findings here are noteworthy in several ways. First, we show that the current distribution of most transboundary species have already emerged from the historical record and that virtually all EEZs have seen at least one emerging species since 2006 (Figure \ref{fig:Figure5}). Second, we identify regional patterns of complex changes in SSR and levels of vulnerability of transboundary fisheries management to climate change (Figure \ref{fig:Figure3}). Third, we identified that the historical share proportion of transboundary species will change by 2030 in most neighboring EEZs (Figure \ref{fig:Figure5}).

#### Emergence characteristics of transboundary species

We set out to evaluate the emergence characteristics of transboundary stocks across the global EEZs. Our results suggest that in most of the EEZs of the world more than half of the transboundary species already emerged since the early 2000s (Figure \ref{fig:Figure2}). This is in alignment with previous research showing changes in the composition of marine catches since the 1970's as evidence of climate change [@Cheung:2013gk]. Indeed, an expansion in the distribution of multiple marine species has been documented in several regions of the world [@Poloczanska:2016kk]. For example, in the early 2000s Humboldt squid (*Dosidicus gigas*) substantially expanded its geographic range reaching Washington waters where a new fishery rapidly developed [@Pinsky:2014by; @Zeidberg:2007gs]. In the North East Atlantic, the emergence of Atlantic mackerel (*Scomber scombrus*) in Icelandic waters in 2010, led to years of unsettled international conflict between multiple fishing nations [@Spijkers:2017ij]. The large amount of transboundary species emerging since early 2000s identified in this study can be partially attributed to the time of emergence of several environmental variables (e.g., temperature, oxygen) that influence the distribution of marine species [@Rodgers:2015jq; @Keller:2014jv; @Frolicher:2016dg; @Mahlstein:2011fy]. It has been estimated that since the early 2000s, SST and oxygen ($O_2$) have emerged in 79% and 23% of the global ocean, respectively, and by 2080 SST will have emerged in 90% of the global ocean while ($O_2$) will have doubled under high greenhouse gas emission scenario [@Rodgers:2015jq].
